---
title: "ggpicrust2"
author: "Jochum, Michael D."
format: 
  html:
    theme: sandstone
editor: visual
---

## Objective

The purpose of this script is to integrate the annotated KEGG pathways from the KO outputs from picrust2 and then compare differences observed between Type 2 diabetic and control cohorts using DESEQ2 wrapped inside ggpicrust2.

## Analysis

1.  Data Importation

2.  Data Wrangling

3.  Data Filtering

    ##### 1. Data Import

    Lets begin by loading our libraries and importing our metadata and KO predictions, then use ggpicrust2 to translate the KO to KEGG pathways

```{r}
library(readr)
library(ggpicrust2)
library(tibble)
library(tidyverse)
library(ggprism)
library(patchwork)
library(mosaic)
library(ggh4x)
getwd()

metadata <- read_delim("IRB_Human_Metadata.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) 
metadata
# Load KEGG pathway abundance
kegg_abundance <- ko2kegg_abundance("picrust2/KO_metagenome_out/pred_metagenome_unstrat.tsv") 

```

##### differential pathway analysis

Using the kegg_abundance and metadata, run DESeq2 to obtain the sig. KO assoc. with Type 2 diabetes

```{r}
deseq2_daa_results_df<- pathway_daa(abundance = kegg_abundance, 
                                   metadata = metadata, 
                                   group = "Diabetes_Status",
                                   p.adjust ="BH",
                                   daa_method = "DESeq2",
                                   select = NULL,
                                   reference = "HC") 
deseq2_daa_results_df<-deseq2_daa_results_df%>%filter(!is.na(p_values)) #14

deseq2_daa_results_df%>%filter(p_values<0.05) #14
deseq2_daa_results_df%>%filter(p_adjust<0.05) #4
```

Annotate the significant KO that have annotations

```{r}

# Annotate pathway results using KO to KEGG conversion
deseq2_daa_annotated_sub_method_results_df <- pathway_annotation(pathway = "KO", daa_results_df = deseq2_daa_results_df, ko_to_kegg = TRUE,)
deseq2_daa_annotated_sub_method_results_df


```

Generate the pathway error plot

```{r}

# Generate pathway error bar plot
# Please change Group to metadata$your_group_column if you are not using example dataset
p <- pathway_errorbar(abundance = kegg_abundance, 
                     daa_results_df = deseq2_daa_annotated_sub_method_results_df, 
                     Group = metadata$Diabetes_Status, 
                     p_values_threshold = 0.25, 
                     order = "pathway_class", 
                     select = NULL, 
                     ko_to_kegg = T, 
                     p_value_bar = TRUE, 
                     colors = NULL, 
                     x_lab = "pathway_name")
p
```
